---
title: "ENVS-193DS Final"
author: "Gabriela Davidson"
date: 2026-06-09
format: pdf
---

# Link to Git Hub
[GitHub Repository](https://github.com/gdavidson533/ENVS-193DS_spring-2026_final.git)

# Set Up 

```{r}
#| label: Set_up
#| message: false
#| warning: false
library(tidyverse)   # data wrangling and ggplot
library(here)        # file paths
library(janitor)     # data cleaning
library(DHARMa)      # diagnostic checks for models
library(MuMIn)       # AIC model selection
library(ggeffects)   # model predictions
library(flextable)   # making tables
```

```{r}
nest_data <- read_csv(here("data", "nest_data_final.csv"))
```


## Problem 1. Research writing

### a. Transparent statistical methods

In part 1, they used a logistic regression, since the response variable (probability of American Avocet micro habitat use) is binary and distance to water edge is a continuous predictor. In part 2, they used a chi-square test of independence, since they are testing the relationship between two categorical variables (bird species and habitat type).

### b. Suggestions for rewriting

Part 1: Distance to water edge was a significant predictor of American Avocet microhabitat use, meaning that where a bird was found was related to how far it was from the water's edge.

Part 2: The three bird species — American Avocet, Forster's Tern, and Black-necked Stilt — were not evenly distributed across habitat types (managed pond, non-tidal marsh, salt pond, tidal marsh, and tidal mudflat), suggesting that each species shows preferences for particular habitats.

### c. Further analysis

Presence of wet bare ground (categorical: present or absent, assessed visually during point count surveys) could influence the probability of American Avocet microhabitat use because, according to Schacter et al. (2023), Avocets fed primarily on wet bare ground in 65% of foraging observations, suggesting they strongly select for this substrate type when choosing where to forage. Sites lacking wet bare ground would therefore be expected to have lower probability of Avocet microhabitat use.

## Problem 2. Data analysis


```{r}
#| label: clean_data

nests_clean <- nest_data |> 
  # select only the 3 columns of interest
  select(case_control, height_cm, ant_species) |> 
  # filter to only the 3 ant species of interest
  filter(ant_species %in% c("AB", "RRB", "BBR")) |> 
  # create column with full scientific names
  mutate(ant_species_full = case_when(
    ant_species == "AB"  ~ "Crematogaster sjostedti",
    ant_species == "RRB" ~ "Crematogaster mimosae",
    ant_species == "BBR" ~ "Crematogaster nigriceps"
  )) |> 
  # reorder factor levels: C. sjostedti, C. mimosae, C. nigriceps
  mutate(ant_species_full = fct_relevel(ant_species_full,
                                        "Crematogaster sjostedti",
                                        "Crematogaster mimosae",
                                        "Crematogaster nigriceps"))

# display structure of nests_clean
str(nests_clean)

# display 10 random rows
slice_sample(nests_clean, n = 10)
```


### b. Response variable

In this dataset, a 1 indicates that the tree was a "used" tree, meaning a bird nest of 
any species was found in that tree. A 0 indicates that the tree was an "available" tree, 
meaning it was one of the four nearest neighbors to a nest tree but did not itself 
contain a nest.

### c. Purpose of study

**Tree height** is a continuous variable measured in centimeters. As tree height 
increases, the probability of nest occurrence would increase because taller trees 
may offer greater structural complexity and concealment from predators, making them 
more suitable nest sites; the authors found that for each 1 m increase in height, 
the odds that birds nested in a tree increased by 20%. This information was found in 
the Methods section (paragraph 2) and the Results section (paragraph 1).

**Acacia ant species** is a categorical variable with three levels: *Crematogaster 
sjostedti*, *Crematogaster mimosae*, and *Crematogaster nigriceps*. As ant species 
identity shifts from less aggressive defenders (*C. sjostedti*) to more aggressive 
defenders (*C. mimosae* and *C. nigriceps*), the probability of nest occurrence would 
increase because more aggressive ant species vigorously defend host trees against 
mammalian herbivores and predators, which may also protect bird nests from disturbance 
and predation. This information was found in the Introduction (paragraphs 2–3) and 
the Discussion (paragraphs 1–2).s

